---
title: "Coding Standards"
output: html_document
---

1. Always use a plain text file!
2. Indent code!
3. Limit text by 80 columns of text

In RStudio: Preferences -> code -> tab width
then CMD+A -> CMD+I
Best use 8 indent spaces

4. Limit the length of individual functions. Limit to one single action.

#Markdown

* *italics*

* ##Heading 2

* ###Heading 3

- first item
- second item
- third item

1. first item
2. second item
3. third item

The numbers in a numbered list do not have to nessesserally be in order!

[link to google](www.google.com)

This is an example of a reference [1]

New lines in text  
require double space at the end of the line

[1]: www.google.com "google"

#R Markdown

* Integration of R in Markdown (duh!).  
* Created by George Gruber and Aaron Swarz  
* A simplified version of markjup languages
* Allows one to focus on writing rather than formatting
* Miniam intuitive formatting elements
* Easily converted to plain HTML
* Allows one to create "live" code with R
* R code is processed during the evaluation of the markdown
* Results from R (plots and other stuff) get inserted into the HTML  
Hence, R Markdown is a **literate statistical programming tool**  
`.Rmd` extension  
One major thing: **if the document exits, it means that the underlying R code has to have worked**!

* R Markdown can be converted to html via `knitr`
* Can also be converted to HTML with `markdown` package
* Can be easily managed in RStudio (but not required)
* `slidify` package allows to create slides from R code

```{r install slidify}
if(!require("slidify")){
  if(!require("devtools")){
    install.packages("devtools")
    library("devtools")
  }
  install_github('slidify', 'ramnathv')
  library("slidify")
}
```

#knitr

##Part 2

`knitr` is good for

* Manuals
* Short to medium length techinical documents
* Tuturials
* Reports (especially generated regularly)
* Data processing and document summaries  

`knitr` is **not** good for  

* Very long research articles
* Complex and time-consuming computations (sometimes you can't really do much about it)
* Documents that require precise formatting

##Part 3

A more sophisticated way to use `knitr`

```{r}
if(!require("knitr")){
  install.packages("knitr")
  library("knitr")
}
#knit2html("coding.standards.Rmd")
#browseURL("coding.standards.html")
```
`echo=FALSE` makes the code block silent, returning no output.  
To hide the results use `results="hide"`.  
Embedding a function: `r "hi"`, this is a random number from uniform distribution `r runif(1)`.  
`fig.height=4` tells how tall should be an output from `plot()`. The image will be embedded in the html as a long 
string of seemingly random letters. Same for the width.  

```{r xtable}
if(!require("xtable")){
  install.packages("xtable")
  library("xtable")
}
```

`xtable` allows to convert r objects to htms or latex directly.

`results="asis"` 

##Global options

```{r setoptions, echo=FALSE}
opts_chunk$set(echo = FALSE, results = "hide")

```

`{r cache=TRUE}` is very usefull when the computation takes a long time to process. The results get stored on the disk and loaded later. One caveat is that if one chunk depends on the previous one, the link might brake. Not entirely everyghin is *cacheable*.






