---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
##*Quick preparation*

Loading libraries:

```{r loading libraries}
if(!require("plyr")){
  install.packages("plyr")
  library("plyr")
}
if(!require("dplyr")){
  install.packages("dplyr")
  library("dplyr")
}
if(!require("tidyr")){
  install.packages("tidyr")
  library("tidyr")
}
if(!require("knitr")){
  install.packages("knitr")
  library("knitr")
}
if(!require("ggplot2")){
  install.packages("ggplot2")
  library("ggplot2")
}
```

Setting options:

```{r setoptions}
opts_chunk$set(echo = TRUE,fig.width = 7, fig.height = 7)

```

## Loading and preprocessing the data

```{r loading data}
#read the data from the zip archive
mov.data<-read.csv(file = unz("activity.zip", "activity.csv" ), stringsAsFactors = FALSE, header = TRUE)
#convert the data to data (duh!) from character
mov.data<-mov.data %>% transform(.,date=as.Date(x=date))
#quick sanity check
head(mov.data)
#The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
assertthat::are_equal(nrow(mov.data),17568)
```


## What is mean total number of steps taken per day?

```{r mean total steps}
#group by date, remove NAs, plot as histogram
summary.sum<-mov.data %>% group_by(date) %>% 
  summarize(sum.steps=sum(steps, na.rm = TRUE))
summary.sum %>% ggplot(data=., mapping=aes(x=sum.steps)) + geom_histogram(binwidth=1000, fill="steelblue", color="black") +
  xlab("number of steps per day") + ggtitle("Total Number Of Steps Per Day")
```

Mean steps per day: **`r as.integer(mean(summary.sum$sum.steps))`**, median is: **`r median(summary.sum$sum.steps)`**

## What is the average daily activity pattern?

```{r average daily activity pattern}
#group by interval and get the mean steps per interval
summary.mean <- mov.data %>% group_by(interval) %>% summarize(mean.steps=mean(steps, na.rm = TRUE))
summary.mean %>% ggplot(data=., mapping=aes(x=interval, y=mean.steps))+geom_line(color="steelblue") +
  xlab("interval (5 mins)") + ylab("mean steps per interval")
```

Maximum number of steps per 5 minute interval: **`r as.integer(max(summary.mean$mean.steps, na.rm=TRUE))`**, which corresponds to the interval: **`r summary.mean[which(summary.mean$mean.steps==max(summary.mean$mean.steps, na.rm=TRUE)),1]`**

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
