---
title: 'Tornados and Floods: Two Main Causes of Damage to Health and Ecnonomy of the United States, according to NOAA'
output: html_document
---

####*by Alex Tuzhikov, August 15 2015*

##Quick preparation:
Loading libraries:

```{r loading libraries}
if(!require("data.table")){
  install.packages("data.table")
  library("data.table")
}
if(!require("plyr")){
  install.packages("plyr")
  library("plyr")
}
if(!require("dplyr")){
  install.packages("dplyr")
  library("dplyr")
}
if(!require("tidyr")){
  install.packages("tidyr")
  library("tidyr")
}
if(!require("knitr")){
  install.packages("knitr")
  library("knitr")
}
if(!require("ggplot2")){
  install.packages("ggplot2")
  library("ggplot2")
}
if(!require("lubridate")){
  install.packages("lubridate")
  library("lubridate")
}
if(!require("R.utils")){
  install.packages("R.utils")
  library("R.utils")
}
if(!require("rgeos")){
  install.packages("rgeos")
  library("rgeos")
}
if(!require("rgdal")){
  install.packages("rgdal")
  library("rgdal")
}
if(!require("readr")){
  install.packages("readr")
  library("readr")
}
if(!require("maptools")){
  install.packages("maptools")
  library("maptools")
}
if(!require("mapproj")){
  install.packages("mapproj")
  library("mapproj")
}
if(!require("gridExtra")){
  install.packages("gridExtra")
  library("gridExtra")
}
```

Setting options:

```{r setoptions}
opts_chunk$set(echo = TRUE,fig.width = 7, fig.height = 7)

```

#Synopsis
some blablah goes here...

##Data processing

###Obtaining the data  

```{r data download, cache=TRUE}
#check if the file has alerady been downloaded
storm.data<-"StormData"
storm.data.csv<-paste0(storm.data,".csv")
storm.data.bz<-paste0(storm.data.csv,".bz2")

if(!file.exists(storm.data.bz)){
  #download the file
  download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                destfile=storm.data.bz, method = "curl")
}
#extract the data and select only the columns that may be valuable for this study
noaa.data<- read.csv(file=storm.data.bz, header=TRUE, stringsAsFactors = FALSE) %>% 
  select(STATE,COUNTYNAME,BGN_DATE,EVTYPE,MAG,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```

Apparently, some of the data need to get tidy. See below:

```{r}
head(noaa.data)
```

Here `BGN_DATE` should probably be of a more appropriate class, that represents a date.

```{r cleaning dates}
#simply convert the dates and get rid of the old BGN_DATE column
noaa.data <- noaa.data %>% mutate(EVENT_DATE=mdy_hms(BGN_DATE)) %>% select(-c(BGN_DATE))
```

Now, there is still an issue with the `PROPDMG` value (property damage). Currently,

```{r PROPDMG PROPDMGEXP demo}
unique(noaa.data$PROPDMGEXP)
```

and we can not use these values "as is", so some conversion is due here as well. I will assign a numeric value, which corresponds to some power of ten to each of the values:

```{r PROPDMG PROPDMGEXP fix}
#            "K" "M" ""  "B" "m" "+" "0" "5" "6" "?" "4" "2" "3" "h" "7" "H" "-" "1" "8"
power.val<-c( 3,  6,  0,  9,  6,  0,  0,  5,  6,  0,  4,  2,  3,  0,  7,  2,  0,  1,  8)
val.map<-mapvalues(x = noaa.data$PROPDMGEXP,
          from=c("K","M","", "B","m","+","0","5","6","?","4","2","3","h","7","H","-","1","8"),
          to=sapply(power.val,function(i){return(10^i)}))
rm(power.val)
noaa.data<- noaa.data %>% mutate(PROPDMG=PROPDMG*as.numeric(val.map)) %>% select(-c(PROPDMGEXP))
rm(val.map)
```

Now, the same thing occurs with the `CROPDMG` and `CROPDMGEXP`:

```{r CROPDMG CROPDMGEXP demo and fix}
unique(noaa.data$CROPDMGEXP)
#           ""  "M" "K" "m" "B" "?" "0" "k" "2"
power.val<-c(0,  6,  3,  6,  9,  1,  1,  3,  2)
val.map<-mapvalues(x = noaa.data$CROPDMGEXP,
          from=c("","M","K","m","B","?","0","k","2"),
          to=sapply(power.val,function(i){return(10^i)}))
rm(power.val)
noaa.data<- noaa.data %>% mutate(CROPDMG=CROPDMG*as.numeric(val.map)) %>% select(-c(CROPDMGEXP))
rm(val.map)
```

##Results

Now that the data may be considered clean, let's try to answer the first question:  
**Across the United States, which types of events are most harmful with respect to population health?**  
To do so, let's do a quick summary:

```{r summarize data}
noaa.data.summary<- noaa.data %>% group_by(EVTYPE) %>% 
  summarize(total.fatalities=sum(FATALITIES), total.injuries=sum(INJURIES), 
            total.prop.dmg=sum(PROPDMG), total.crop.dmg=sum(CROPDMG)) %>%
  arrange(-(total.fatalities+total.injuries+total.prop.dmg+total.crop.dmg))
#quick summary
head(noaa.data.summary)
```

Apparently, there is a lot of weather conditions that cause some damage, but only some have devastating effect. To comprehend the information it is best if we take just the top part of the list. The distribution of total damage, caused by the disasters, sorted descending shows that we can safely. For the future it makes sense to use just the top 10.

Let's see which of the natural disasters affect health the most:

```{r health, fig.width = 14}
#let's assume total damage to the health of the popultaion to be a the total fatalities and injuries combined
noaa.data.summary.health<- noaa.data.summary %>% mutate(total.health.dmg=total.fatalities+total.injuries) %>%
  arrange(-total.health.dmg) %>% do({
    .$EVTYPE<-factor(.$EVTYPE, levels=.$EVTYPE)
    return(.[1:10,])
  })
#plot
gp.health<-ggplot(data=noaa.data.summary.health, mapping=aes(x=EVTYPE, y=total.health.dmg))+geom_bar(stat="identity",fill="red") +
  theme(axis.text.x=element_text(angle=90),legend.position="none",
        strip.background=element_blank(),
        plot.background=element_rect(fill = "white", colour = "white"),
         panel.background = element_blank())+
  ylab("Total Health Damage Cases")+xlab("Disaster Type")
#map the events
#get summary by state
noaa.data.health.by.state<- noaa.data %>% group_by(STATE, EVTYPE) %>% 
  summarize(total.health.dmg=sum(FATALITIES+INJURIES)) %>% filter(EVTYPE %in% noaa.data.summary.health$EVTYPE)
#select only the top ones
evtype<-c("TORNADO", "EXCESSIVE HEAT", "FLOOD", "LIGHTNING")
noaa.data.health.by.state.plot<- noaa.data.health.by.state %>% 
  filter(EVTYPE %in% evtype) %>% 
  mutate(EVTYPE=factor(EVTYPE, levels = evtype))

#prepare the hexagons
us.states.geo<-"us_states_hexgrid.geojson"
if(!file.exists(us.states.geo)){
  download.file("https://gist.githubusercontent.com/hrbrmstr/51f961198f65509ad863/raw/219173f69979f663aa9192fbe3e115ebd357ca9f/us_states_hexgrid.geojson", 
              us.states.geo, method = "curl")
}
state.hex<- readOGR(us.states.geo, "OGRGeoJSON")
centers <- cbind.data.frame(data.frame(gCentroid(state.hex, byid=TRUE), id=state.hex@data$iso3166_2))
#map to the US map
us.map <- fortify(state.hex, region="iso3166_2")
noaa.data.health.by.state.plot <- noaa.data.health.by.state.plot %>% filter(STATE %in% unique(us.map$id))
#plot
us.map.health.gp<-ggplot()+
  geom_map(data=us.map, map=us.map, mapping=aes(x=long, y=lat, map_id=id), color="white", size=0.5)+
  #now the actual layer
  geom_map(data=noaa.data.health.by.state.plot, map=us.map,
                    aes(fill=total.health.dmg, map_id=STATE))+
  geom_map(data=noaa.data.health.by.state.plot, map=us.map, aes(map_id=STATE), fill="white", alpha=0, color="white", show_guide=FALSE)+
  geom_text(data=centers, aes(label=id, x=x, y=y), color="white", size=3)+
  scale_fill_distiller(name="Damages to Health\nand\nCasualties", palette="Reds", 
                       na.value="white", labels=sprintf("%d%%", seq(0,100,25)))+
  coord_map()+facet_wrap(~EVTYPE,nrow = 2)+
  theme(panel.border=element_blank(),
        plot.title=element_text(face="bold", size=24),
        panel.grid=element_blank(),
        axis.ticks=element_blank(),
        axis.text=element_blank(),
        strip.background=element_blank(),
        plot.background=element_rect(fill = "white", colour = "white"),
        strip.text=element_text(face="bold", hjust=0, size=14),
        strip.background=element_rect(colour="white"),
        legend.position="bottom",
        panel.background = element_blank(),
        legend.title.align=1)+
  xlab("")+ylab("")

grid.arrange(gp.health, us.map.health.gp, ncol=2,widths=c(1,3))
```
So, apparently, tornados keep harasing mid-west, while it can get really hot in Montana (surprising actually) and Texas is being flooded badly sometimes, Florida has lightning issues. Unfortunately, the most devastating disaster, tornado, is wastly spread according to the map and have the greates impact on the national health overall.

That being said, let's see for the second question:  
**Across the United States, which types of events have the greatest economic consequences?**

To answer this question lets look at the damages to the property and crops.

```{r economy, fig.width = 14}
#first the summary to see which variables are we looking for
noaa.data.summary.economy<- noaa.data.summary %>% mutate(total.econ.dmg=total.prop.dmg+total.crop.dmg) %>%
  arrange(-total.econ.dmg) %>% do({
    .$EVTYPE<-factor(.$EVTYPE, levels=.$EVTYPE)
    return(.[1:10,])
  })
#most damagind factors
evtype<-c("FLOOD", "HURRICANE/TYPHOON", "TORNADO", "STORM SURGE")
#get summary by state
noaa.data.economy.by.state<- noaa.data %>% group_by(STATE, EVTYPE) %>% 
  summarize(total.econ.dmg=sum(PROPDMG+CROPDMG)) %>% filter(EVTYPE %in% noaa.data.summary.economy$EVTYPE)
noaa.data.economy.by.state.plot<- noaa.data.economy.by.state %>% 
  filter(EVTYPE %in% evtype) %>% 
  mutate(EVTYPE=factor(EVTYPE, levels = evtype))
#plot the damage level
gp.economy<-ggplot(data=noaa.data.summary.economy, mapping=aes(x=EVTYPE, y=total.econ.dmg))+
  geom_bar(stat="identity",fill="chartreuse4") +
  theme(axis.text.x=element_text(angle=90),legend.position="none",
        strip.background=element_blank(),
        plot.background=element_rect(fill = "white", colour = "white"),
         panel.background = element_blank())+
  ylab("Total Economy Damage Cases")+xlab("Disaster Type")
#plot maps
us.map.economy.gp<-ggplot()+
  geom_map(data=us.map, map=us.map, mapping=aes(x=long, y=lat, map_id=id), color="white", size=0.5)+
  #now the actual layer
  geom_map(data=noaa.data.economy.by.state.plot, map=us.map,
                    aes(fill=total.econ.dmg, map_id=STATE))+
  geom_map(data=noaa.data.economy.by.state.plot, map=us.map, aes(map_id=STATE), fill="white", alpha=0, color="white", show_guide=FALSE)+
  geom_text(data=centers, aes(label=id, x=x, y=y), color="white", size=3)+
  scale_fill_distiller(name="Damages to Property\nand\nEconomy", palette="Greens", 
                       na.value="white", labels=sprintf("%d%%", seq(0,100,25)))+
  coord_map()+facet_wrap(~EVTYPE,nrow = 2)+
  theme(panel.border=element_blank(),
        plot.title=element_text(face="bold", size=24),
        panel.grid=element_blank(),
        axis.ticks=element_blank(),
        axis.text=element_blank(),
        strip.background=element_blank(),
        plot.background=element_rect(fill = "white", colour = "white"),
        strip.text=element_text(face="bold", hjust=0, size=14),
        strip.background=element_rect(colour="white"),
        legend.position="bottom",
        panel.background = element_blank(),
        legend.title.align=1)+
  xlab("")+ylab("")

grid.arrange(gp.economy, us.map.economy.gp, ncol=2,widths=c(1,3))
```

According to the plot above, the most damaging to the US economy are floods and typhoons, that are mainly occuring in California and Florida respectively.





